Title Goes Here

### 2021-03-04 analysis/paper/paper.Rmd
Draft resultados

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 41,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

fig.class <- list(
    pagewidth = list(
        fig.width = page.width/25.4,
        out.width = paste0(page.width, "mm"),
        # fig.align = "center",
        fig.env = "figure*",
        out.extra = " "
    )
)

@@

theme_set(theme_shceof(base_size = 16))

@@

guide_colorsteps_bottom <- function(width = 25, height = 0.5, even.steps = FALSE, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

pattern_dots <- ggplot2::ggproto("GeomDots", ggpattern::GeomPolygonPattern)
ggplot2::update_geom_defaults(pattern_dots, 
                              list(pattern = "circle",
                                   colour = NA,
                                   pattern_colour = "black",
                                   pattern_fill = "black",
                                   pattern_density = 0.1,
                                   pattern_alpha = 1,
                                   fill = NA,
                                   pattern_spacing = 0.02))

@@

We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each spring  and computing the Fourier spectrum.

@@

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.

@@

Another characteristic seen in Figure \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between temporal series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

@@

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`. This also nearly minimises the correlation with the imaginary part (\@ref(fig:rotations-pc1)).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (\@ref(fig:rotations-pc2)). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

@@

@@

```{r ceofs-1, fig.cap = "Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Spatial patterns. Real part in shading, imaginary part in contours"}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Temproal series."}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

@@

(ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.

```{r extended-series, fig.cap = "(ref:extended-series-cap)"}
pattern <- ceof$eof[[1]]$right %>%
  sep_ReIm() %>%
  setnames("hgt", "EOF")

@@

# Check that the extended index coincides with the
# original index.
cors <- era_more %>%
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>%
  na.omit() %>%
  .[, cor(hgt, i.hgt), by = .(PC, part)]

@@

Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis). There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`). There is no significant trend in any of the complex parts of EOF2.

@@

  if ("lev" %in% colnames(data)) {
    data <- data[lev %in% which_levs]
  } else {
    data[, lev := which_levs]
  }
  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
        geom_contour_tanaka() +
        scale_fill_divergent(var_title, guide = guide_colorstrip_bottom())
    ) +
    geom_contour_pval(aes(z = p.value)) +

@@

    scale_x_longitude() +
    scale_y_latitude() +
    facet +
    coord_polar() +
    axis_labs_smol +
    tag_facets("rc")
}

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt") +
  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
  # scale_linetype(guide = "none") +
  NULL
```

@@

```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)"}
plot_regr(regr, "PC2", c(200, 50), "hgt") 
```

@@

```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)"}
plot_regr(regr, "PC1", c(200, 50), "t")
```

@@

```{r t-vertical, fig.cap = "Regression of EOF1 with mean air temperature between 80ºS and 25ºS."}
t %>% 
  .[lat %between% c(-80, -35)] %>% 
  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  .[, t := Anomaly(t), by = .(time, lev)] %>% 
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
  ggplot(aes(lon, lev)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), 
                      breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(25)) + 
  scale_y_level() +
  scale_x_longitude() +
  facet_grid(~term) +
  axis_labs_smol
```

@@

```{r o3-regr, fig.cap = "Regression of standarised EOF1 and Total Ozone column."}
o3_regr %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   guide = guide_colorsteps_bottom()) + 
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  facet_grid(PC~term) +
  axis_labs_smol +
  tag_facets("panel")
```

@@

```{r wave1-o3, fig.cap = "EOF1 and wave1 o3 amplitude."}
wave1_o3 <- o3 %>% 
  .[lat %between% o3wave_lats] %>% 
  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(time, lon)] %>% 
  .[, FitWave(toc, 1), by = .(time)]

@@

temp %>% 
  ggplot(aes(Mod(hgt), amplitude)) +
  scale_x_continuous("EOF1 Amplitude") +
  scale_y_continuous("O3 Wave-1 Amplitude") +

temp %>% 
  copy() %>%
  na.omit() %>% 
  qwrap(phase = c(0, 2*pi) ~ c(-pi, pi)) %>% 
  ggplot(aes(Arg(hgt), phase)) +
  scale_x_continuous("EOF1 Phase",
                     breaks = seq(-pi, pi, by = 15*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("O3 Wave-1 Phase",
                     breaks = seq(-pi, pi, by = 15*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
  plot_layout(ncol  = 1) &
  geom_point()  &
  geom_smooth(method = "lm") 
```

@@

```{r psa-eof2, fig.cap = "EFO2 and PSA"}
pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
  copy() %>% 
  .[, PC := NULL] %>% 
  .[]

@@

pc2[psa_time, on = "time"] %>% 
  .[PC != "SAM"] %>% 
  copy() %>% 
  # .[, hgt := rotate(hgt, psa_angle)] %>%
  sep_ReIm() %>% 
  ggplot(aes(value, hgt)) +
  geom_point() +
  scale_y_continuous("EOF2") +
  scale_x_continuous("PSA") +
  facet_grid(part ~ PC) +
  panel_background +
  tag_facets("rc")
```

@@

```{r phase-histogram, fig.cap = "Histogram of phase distributions of EOF2."}
cut(ceof$eof[[1]], 2)$left %>% 
  .[, arg := Arg(hgt)] %>% 
  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
  ggplot(aes(arg)) +
  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
  annotate("text",
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
           color = "black", size = 6,
           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
  geom_rug() +
  scale_y_continuous("Number of years") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
```

@@

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
time <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)]

@@

sam_r2 %>%
  ggplot(aes(lev, estimate^2)) +
  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²", limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc") +
  panel_background
```

@@

```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale."}
eof_filtered <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1, data = .)

@@

eof_filtered %>% 
  .$left %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  scale_fill_divergent(guide = "none") +
  geom_qmap(~.x[lat <= -20]) +  
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  axis_labs_smol 
```

@@

```{r fake-eof, fig.cap = "Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude."}
pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

@@

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)"}
plot_sst_psi(which_pc = "PC2", which_lev = 200)
```

@@

```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)"}
plot_sst_psi(which_pc = "PC1", which_lev = 200)
```

@@

# Appendix

## Chosen rotations of the EOFs

>>>>

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 42,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

fig.class <- list(
  colwidth = list(
    fig.width = 5
  ),
    pagewidth = list(
        fig.width = page.width/25.4,
        out.width = paste0(page.width, "mm"),
        # fig.align = "center",
        fig.env = "figure*",
        out.extra = " "
    )
)

@@

theme_set(theme_shceof(base_size = 12))

@@

guide_colorsteps_bottom <- function(width = 20, height = 0.5, even.steps = FALSE, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

pattern_dots <- ggplot2::ggproto("GeomDots", ggpattern::GeomPolygonPattern)
ggplot2::update_geom_defaults(pattern_dots, 
                              list(pattern = "circle",
                                   colour = NA,
                                   pattern_colour = "black",
                                   pattern_fill = "black",
                                   pattern_density = 0.05,
                                   pattern_alpha = 1,
                                   fill = NA,
                                   pattern_spacing = 0.025))

update_geom_defaults(GeomSmooth, list(colour = "#0d52bf"))

@@

We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between `r knitr::combine_words(LatLabel(o3wave_lats))` for each Spring and computing the Fourier spectrum.

@@

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units). The numbers at the top-left of each panel ate the variance explained by each EOF.

@@

Another characteristic seen in Figure \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

@@

@@

(ref:rotations-cap) a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of  Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` and b) $R^2$ between the Real and Imaginary EOF2 and the ONI index for different values of rotation. The vertical line marks the selected rotation, which maximises de correlation.

```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=5, fig.height=2.5}
r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
    paste("2")
}, "", ""))

rotations_o3 %>%
  ggplot(aes(rotation, cor)) +
  geom_vline(xintercept = best_rotation_pc1) +
  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
  geom_line(aes(color = part)) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 60*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("Correlation") +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +

rotations_pc2 %>%
  melt(id.vars = "rotation") %>%
  ggplot(aes(rotation, value^2)) +
  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
  geom_vline(xintercept = best_rotation_pc2) +
  geom_line(aes(color = variable)) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 60*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous(r2) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +

  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`. This also nearly minimises the correlation with the imaginary part (Figure \@ref(fig:rotations)a).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure \@ref(fig:rotations)b). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

@@

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Real part in shading, imaginary part in contours", fig.width=5, fig.height=5}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=4, fig.width = 5}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

@@

```{r era_more}
pattern <- ceof$eof[[1]]$right %>%
  sep_ReIm() %>%
  setnames("hgt", "EOF")

@@

(ref:extended-series-cap) Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.

```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=4, fig.width = 5}
# Check that the extended index coincides with the
# original index.
cors <- era_more %>%
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>%
  na.omit() %>%
  .[, cor(hgt, i.hgt), by = .(PC, part)]

@@

Figure \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis). There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`). There is no significant trend in any of the complex parts of EOF2.

@@

  if ("lev" %in% colnames(data)) {
    data <- data[lev %in% which_levs]
  } else {
    data[, lev := which_levs]
  }

  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
    scale_fill_divergent(var_title, breaks = AnchorBreaks(0, exclude = 0),
                         guide = guide_colorstrip_bottom(20)) +
    geom_contour_pval(aes(z = p.value)) +

@@

    scale_x_longitude() +
    scale_y_latitude() +
    facet +
    coord_polar() +
    axis_labs_smol +
    tag_facets("rc")
}
```

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=5.4, fig.width=5}
plot_regr(regr, "PC1", c(200, 50), "hgt") +
  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
  # scale_linetype(guide = "none") +
  NULL

@@

@@

```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)", fig.height=5.4, fig.width=5}
plot_regr(regr, "PC2", c(200, 50), "hgt")

@@

```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)",  fig.height=5.4, fig.width=5}
plot_regr(regr, "PC1", c(200, 50), "t")

@@

(ref:t-vertical-cap) Regression of EOF1 with mean air temperature averaged between 80ºS and 25ºS for the period 1979 -- 2019.

```{r t-vertical, fig.cap = "(ref:t-vertical-cap)", fig.height=3.5, fig.width=5}
t %>% 
  .[lat %between% c(-80, -35)] %>% 
  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  .[, t := Anomaly(t), by = .(time, lev)] %>% 
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
  ggplot(aes(lon, lev)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), 
                      breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(20)) + 
  scale_y_level() +
  scale_x_longitude(ticks = 60) +
  facet_grid(~term) +
  axis_labs_smol
```

@@

(raf:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column for the 1979 -- 2019 period.

```{r o3-regr, fig.cap = "(raf:o3-regr-cap)", fig.height=4, fig.width = 5}
o3_regr %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   guide = guide_colorsteps_bottom(20)) + 
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  facet_grid(PC~term) +
  axis_labs_smol +
  tag_facets("panel")

@@

(ref:wave1-o3-cap) Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between `r knitr::combine_words(LatLabel(o3wave_lats))` for the period 1979 -- 2019.

```{r wave1-o3, fig.cap = "(ref:wave1-o3-cap)", fig.width=5, fig.height=3}
wave1_o3 <- o3 %>% 
  .[lat %between% o3wave_lats] %>% 
  .[, .(toc = mean(toc, na.rm = TRUE)), by = .(time, lon)] %>% 
  .[, FitWave(toc, 1), by = .(time)]

@@

temp %>% 
  ggplot(aes(Mod(hgt), amplitude)) +
  scale_x_continuous("EOF1 Amplitude") +
  scale_y_continuous("O3 Wave-1\nAmplitude") +

temp %>% 
  copy() %>%
  na.omit() %>% 
  qwrap(phase = c(0, 2*pi) ~ c(-pi, pi)) %>% 
  ggplot(aes(Arg(hgt), phase)) +
  scale_x_continuous("EOF1 Phase",
                     breaks = seq(-pi, pi, by = 15*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("O3 Wave-1\nPhase",
                     breaks = seq(-pi, pi, by = 30*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
  plot_layout(ncol  = 1) &
  geom_point()  &
  geom_smooth(method = "lm")

@@

(ref:psa-eof2-cap) Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies [@mo2001]. For the period 1979 -- 2019.

```{r psa-eof2, fig.cap = "(ref:psa-eof2-cap)", fig.width=5, fig.height=4}
pc2 <- ceof$eof[[1]]$left[PC == "PC2"] %>% 
  copy() %>% 
  .[, PC := NULL] %>% 
  .[]

@@

pc2[psa_time, on = "time"] %>% 
  .[PC != "SAM"] %>% 
  copy() %>% 
  # .[, hgt := rotate(hgt, psa_angle)] %>%
  sep_ReIm() %>% 
  ggplot(aes(value, hgt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_smooth(method = "lm") + 
  scale_y_continuous("EOF2") +
  scale_x_continuous("PSA") +
  facet_grid(part ~ PC) +
  panel_background +
  tag_facets("rc")

@@

(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.

```{r phase-histogram, fig.cap = "(ref:phase-histogram-cap)", fig.width=2.5, fig.height=2.5}
cut(ceof$eof[[1]], 2)$left %>% 
  .[, arg := Arg(hgt)] %>% 
  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
  ggplot(aes(arg)) +
  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
  annotate("text",
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
           color = "black", size = 3,
           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
  geom_rug() +
  scale_y_continuous("Number of years") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
```

@@

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)", fig.height=4, fig.width=5}
time <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)]

@@

sam_r2 %>%
  ggplot(aes(lev, estimate^2)) +
  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous(r2, limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc") +
  panel_background
```

@@

```{r eof-filtered, fig.cap = "Spatial pattern of the leading EOF of 200 hPa geopotential height with the variability of EOF2 filtered out South of 20º. Arbitrary scale.", fig.width=2.5, fig.height=2.5}
eof_filtered <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := resid(.lm.fit(cbind(1, Real, Imaginary), hgt)), .(lon, lat, lev, season(time))] %>%
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, lev, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1, data = .)

@@

eof_filtered %>% 
  .$left %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  scale_fill_divergent(guide = "none") +
  geom_qmap(~.x[lat <= -20]) +  
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  theme(axis.text = element_blank())
```

@@

(ref:fake-eof-cap) EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.

```{r fake-eof, fig.cap = "(ref:fake-eof-cap)", fig.width=5, fig.height=2.5}
pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

@@

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.with = 6, fig.height=4}
plot_sst_psi(which_pc = "PC2", which_lev = 200)
```

@@

```{r sst-psi-1, fig.cap = "(ref:sst-psi-1-cap)", fig.with = 6, fig.height=4}
plot_sst_psi(which_pc = "PC1", which_lev = 200)
```

@@
